Final Project Part 3 Report

### 2024-12-06 final_paper/final_report.Rmd
inital

The model's assumptions were validated using residual plots, Q-Q plots, and statistical tests like Cook’s Distance. Linearity and homoscedasticity were assessed through residuals vs. fitted values and residuals vs. predictors plots, while normality of residuals was evaluated with Q-Q plots and histograms, which showed minor deviations in the tails. These diagnostic checks ensured the model's validity and strengthened confidence in its results and inferences.

We used `ggplot2` for visualizations and `car` for diagnostics to explore the relationships between the 13 relavent predictors and the response variable (NHPI). Scatterplots and Variance Inflation Factors (VIF) were used to verify linearity and identify multicollinearity issues among predictors. Exploratory data analysis (EDA) was conducted on both training and test datasets to ensure consistency in variable distributions and to detect any anomalies. Multicollinearity was addressed through transformations, which improved model accuracy by refining variable relationships.

@@

Using the `lm` function in R, an initial regression model was constructed by selecting predictors based on their theoretical relevance and statistical significance. To improve model performance, log transformations were applied to skewed predictors to enhance linearity, and a Box-Cox transformation was used on the response variable to correct residual skewness and stabilize variance. The final model was refined through rigorous diagnostic checks, ensuring it adhered to regression assumptions and achieved optimal accuracy.

@@

After implementing transformations and adjustments, diagnostics were rechecked using residual plots to confirm improvements in linearity, homoscedasticity, and normality. Cook’s Distance was revisited to ensure no new influential points emerged, and Variance Inflation Factor (VIF) values were reassessed to verify acceptable multicollinearity levels. These steps validated the effectiveness of our transformations and actions; if any issues persisted, the process was revisited at Step 3 to refine the model further.

@@

Model performance was assessed using metrics such as $R^2$, Adjusted $R^2$, AIC, BIC, and RMSE. These metrics demonstrated a strong fit, indicating that the model explained the majority of the variability in the response variable. The RMSE highlighted high predictive accuracy, and most predictors were statistically significant, confirming the model's reliability and effectiveness in capturing key relationships.

@@

Model validation was conducted using performance metrics on test data to assess its generalizability and predictive accuracy. Key metrics, such as Mean Squared Error (MSE) and $R^2$, were calculated to evaluate how well the model performed on unseen data. These results confirmed the model's ability to maintain high accuracy and reliability when applied to new datasets.

>>>>

The model's assumptions were validated using residual plots, Q-Q plots, and statistical tests like Cook’s Distance. Linearity and homoscedasticity were assessed through residuals vs. fitted values and residuals vs. predictors plots(Consumer Price Index (CPI), Unabsorbed Homes and Completed Construction Homes), while normality of residuals was evaluated with Q-Q plots and histograms, which showed minor deviations in the tails. These diagnostic checks ensured the model's validity and strengthened confidence in its results and inferences. A further BOX-COX transformation was applied to the response variable (NHPI) to address issues of skewness and heteroscedasticity.

We used `ggplot2` for visualizations and `car` for diagnostics to explore the relationships between the 13 relevant predictors and the response variable (NHPI). Scatterplots and Variance Inflation Factors (VIF) were used to verify linearity and identify multicollinearity issues among all the predictors. Exploratory data analysis (EDA) was conducted on both training and test datasets to ensure consistency in variable distributions and to detect any anomalies. Multicollinearity was addressed through transformations, which improved model accuracy by refining variable relationships.

@@

Using the `lm` function in R, an initial regression model was constructed by selecting predictors based on their theoretical relevance and statistical significance, naming the predictors: CPI, unabsorbed homes, completed construction homes, and absorption. To improve model performance, log transformations were applied to skewed predictors absorption, completed construction homes, and unabsorbed homes, addressing non-linearity and improving their relationships with the response variable (NHPI). A Box-Cox transformation was applied to the New Housing Price Index (NHPI) to correct residual skewness and stabilize variance, ensuring better adherence to regression assumptions. The final model underwent rigorous diagnostic checks, including residual plots, Q-Q plots, and statistical tests like Cook’s Distance, to validate linearity, homoscedasticity, and normality of residuals. These refinements ensured the model not only satisfied regression assumptions but also achieved optimal accuracy and robustness in explaining the variability in NHPI.

@@

After implementing transformations and adjustments, diagnostics were rechecked using residual plots to confirm improvements in linearity, homoscedasticity, and normality. Residuals vs. fitted values and residuals vs. predictors plots were examined to ensure that the transformations effectively addressed patterns of non-linearity and heteroscedasticity. Q-Q plots and histograms of residuals were revisited to evaluate the distribution of residuals, confirming that deviations from normality, particularly in the tails, were significantly reduced. Cook’s Distance was revisited to ensure no new influential points emerged that could unduly affect the regression coefficients, and Variance Inflation Factor (VIF) values were reassessed to verify that multicollinearity levels among predictors, CPI and absorption, were within acceptable thresholds. These iterative steps validated the effectiveness of the transformations and adjustments. If any persistent issues were detected, the process was revisited at Step 3, refining predictor transformations or re-evaluating model specifications to further enhance accuracy and adherence to regression assumptions.

@@

Model performance was assessed using metrics such as $R^2$, Adjusted $R^2$, AIC, BIC, and RMSE. The model demonstrated a strong fit,indicating that the model effectively explained the majority of the variability in the response variable. A high $R^2$ and Adjusted $R^2$ value suggest that a large proportion of the variability in NHPI was captured by the model, with the Adjusted $R^2$ taking into account the number of predictors used, thus avoiding overfitting. The AIC and BIC, which penalize for model complexity, were both relatively low, indicating a good balance between model fit and complexity. The RMSE, which reflects the average prediction error, was also low, suggesting that the model made accurate predictions with minimal deviation from the observed values. Most of the predictors including CPI, completed construction homes, and absorption, were statistically significant, with low p-values confirming strong relationships with the response variable (NHPI), further validating the model’s reliability and its effectiveness to capture the key factors influencing NHPI.

@@

Model validation was conducted using performance metrics on test data to assess its generalizability and predictive accuracy. Key metrics, such as Mean Squared Error (MSE) and $R^2$, were calculated to evaluate how well the model performed on unseen data. A low MSE indicates that the model's predictions are close to the actual values, suggesting high predictive accuracy. Conversely, a high MSE signals large deviations between predicted and observed values, implying that the model struggles to generalize to new data. Similarly, a high $R^2$ indicates that the model explains a large portion of the variability in the response variable, showing strong predictive performance. A low or negative $R^2$, on the other hand, suggests that the model performs poorly, failing to capture the underlying patterns in the test data. These results are critical for understanding how well the model can be expected to perform on future datasets.
